---
title: "Artist Collaboration Network Analysis"
author: "Brandon Emory & Rudy Tjong"
date: "2025-05-28"
output: 
  html_document:
    toc: yes
    toc_depth: 2
    toc_float: true
---

# Summary
In this project, edges represent collaborations—specifically when artists are featured together on a song—while nodes (or vertices) represent individual artists. The data set includes information on approximately 20,000 primary artists whose songs appeared on the Spotify weekly charts, along with around 136,000 additional artists who were featured on songs with at least one charting artist.

The dataset also captures the frequency and structure of these collaborations, allowing us to generate a large-scale network comprising over 135,000 artists as nodes and more than 300,000 edges representing collaborative links between them.

**Data Sources**:
-Aggregated weekly Spotify chart data was collected from [Kworb](https://kworb.net/spotify/)
-Artist and feature data were scraped from the Spotify API.

**Temporal Coverage**:
-Start Date: September 28, 2013
-End Date: October 9, 2022

## Load required libraries
```{r setup, message = FALSE, error = FALSE}
library(tidyverse)
library(igraph)
library(tidygraph)
library(ggraph)
```

# Step 1: Read the data

```{r load data, message= FALSE, error = FALSE}
e <- read.csv("edges.csv", header = TRUE, stringsAsFactors = FALSE)
n <- read.csv("nodes.csv", header = TRUE, stringsAsFactors = FALSE)

head(e)
head(n)
```

# Step 2: Make sure node IDs match
```{r match nodes, message= FALSE, error=FALSE}
n_subset <- n %>%
  rename(id = spotify_id) %>%
  distinct(id, .keep_all = TRUE) %>%
  slice(1:1000)  #for only first 1000 for faster results

valid_ids <- n_subset$id

edges_filtered <- e %>%
  filter(id_0 %in% valid_ids & id_1 %in% valid_ids)

```

# Step 3: Create graph from edge list and join with node data
```{r graph, message=FALSE, error=FALSE}
g <- graph_from_data_frame(d = edges_filtered, vertices = n_subset, directed = FALSE)

plot(g,
     edge.arrow.size = .4,
     edge.arrow.color = "pink",
     vertex.color = "green", 
     vertex.label = NA, 
     vertex.size = 7)

```


Graph shows us a tight group of popular artist in the center meaning they are some of the biggest names. Most artist here are connected to a few people. A few people however, help link the rest together through lots of collaborations.

# Step 4: Basic network metrics
```{r network, message = FALSE, error = FALSE}
V(g)$degree <- degree(g)
V(g)$betweenness <- betweenness(g)
V(g)$pagerank <- page_rank(g)$vector
```

# Step 5: Community detection using Louvain algorithm
```{r community, message = FALSE, error = FALSE}
louvain_clusters <- cluster_louvain(g)
V(g)$community <- louvain_clusters$membership
```

# Step 6: Top artists by centrality
```{r centrality, message = FALSE, error = FALSE}
top_artists <- V(g)[order(-V(g)$degree)][1:10]

top_artists_info <- data.frame(
  name = V(g)$name[top_artists],
  degree = V(g)$degree[top_artists],
  popularity = V(g)$popularity[top_artists],
  followers = V(g)$followers[top_artists]
)

print(top_artists_info)
```


# Conclusion

This analysis highlights the structure and key players in the Spotify artist collaboration network. Using community detection and centrality metrics, we gain insight into how interconnected the music industry is and which artists play central roles in shaping collaboration trends.



