---
title: "Artist Collaboration Network Analysis"
author: "Brandon Emory & Rudy Tjong"
date: "2025-05-28"
output: 
  html_document:
    toc: yes
    toc_depth: 2
    toc_float: true
---

# Summary
In this project, edges represent collaborations—specifically when artists are featured together on a song—while nodes (or vertices) represent individual artists. The data set includes information on approximately 20,000 primary artists whose songs appeared on the Spotify weekly charts, along with around 136,000 additional artists who were featured on songs with at least one charting artist.

The dataset also captures the frequency and structure of these collaborations, allowing us to generate a large-scale network comprising over 135,000 artists as nodes and more than 300,000 edges representing collaborative links between them.

**Data Sources**:
-Aggregated weekly Spotify chart data was collected from [Kworb](https://kworb.net/spotify/)
-Artist and feature data were scraped from the Spotify API.

**Temporal Coverage**:
-Start Date: September 28, 2013
-End Date: October 9, 2022

## Load required libraries
```{r setup, message = FALSE, error = FALSE}
library(tidyverse)
library(igraph)
library(tidygraph)
library(ggraph)
```

# Step 1: Read the data

```{r load data, message= FALSE, error = FALSE}
e <- read.csv("edges.csv", header = TRUE, stringsAsFactors = FALSE)
n <- read.csv("nodes.csv", header = TRUE, stringsAsFactors = FALSE)

head(e)
head(n)
```

# Step 2: Make sure node IDs match
```{r match nodes, message= FALSE, error=FALSE}
n_subset <- n %>%
  rename(id = spotify_id) %>%
  distinct(id, .keep_all = TRUE) %>%
  slice(1:1000)  #for only first 1000 for faster results

valid_ids <- n_subset$id

edges_filtered <- e %>%
  filter(id_0 %in% valid_ids & id_1 %in% valid_ids)

```

# Step 3: Create graph from edge list and join with node data
```{r graph, message=FALSE, error=FALSE}
g <- graph_from_data_frame(d = edges_filtered, vertices = n_subset, directed = FALSE)

plot(g,
     edge.arrow.size = .4,
     edge.arrow.color = "pink",
     vertex.color = "green", 
     vertex.label = NA, 
     vertex.size = 7)

```


Graph shows us a tight group of popular artist in the center meaning they are some of the biggest names. Most artist here are connected to a few people. A few people however, help link the rest together through lots of collaborations.

# Step 4: Basic network metrics
```{r network, message = FALSE, error = FALSE}
V(g)$degree <- degree(g)
V(g)$betweenness <- betweenness(g)
V(g)$pagerank <- page_rank(g)$vector

top_degree <- V(g)[order(-degree)][1:10]
top_degree_df <- data.frame(
  name = V(g)$name[top_degree],
  degree = V(g)$degree[top_degree]
)
print("Top 10 artists by degree:")
print(top_degree_df)

top_betweenness <- V(g)[order(-betweenness)][1:10]
top_betweenness_df <- data.frame(
  name = V(g)$name[top_betweenness],
  betweenness = V(g)$betweenness[top_betweenness]
)
print("Top 10 artists by betweenness:")
print(top_betweenness_df)

top_pagerank <- V(g)[order(-pagerank)][1:10]
top_pagerank_df <- data.frame(
  name = V(g)$name[top_pagerank],
  pagerank = V(g)$pagerank[top_pagerank]
)
print("Top 10 artists by PageRank:")
print(top_pagerank_df)
```
Top degree artist - _The Him_. The top artist here has worked with the most other artists.
They are very active in collaborations. 
Top betweeness artist - _Cardi B_. This artist connects different groups of artists.
They often work with people who don’t usually work together.
Top pagerank artist - _Epik High_. This artist is very influential in the network.
They are connected to other important artists.


# Step 5: Community detection using Louvain algorithm
```{r community, message = FALSE, error = FALSE}
louvain_clusters <- cluster_louvain(g)
V(g)$community <- louvain_clusters$membership

community_df <- data.frame(
  name = V(g)$name,
  community = V(g)$community
)
print("Sample of artists and their community:")
print(head(community_df, 10))
```
We used the Louvain algorithm to find communities in the artist network.
Each community is a group of artists who tend to collaborate with each other more often.
This helps us see how the music industry is organized into smaller groups or genres, and which artists belong to which group.

# Step 6: Top artists by centrality
```{r centrality, message = FALSE, error = FALSE}
top_artists <- V(g)[order(-V(g)$degree)][1:10]

top_artists_info <- data.frame(
  name = V(g)$name[top_artists],
  degree = V(g)$degree[top_artists],
  popularity = V(g)$popularity[top_artists],
  followers = V(g)$followers[top_artists]
)


print(top_artists_info)
```
This table shows the top 10 artists with the most collaborations in the network.
These artists are highly active and connected in the music industry.


# Conclusion

This analysis highlights the structure and key players in the Spotify artist collaboration network. Using community detection and centrality metrics, we gain insight into how interconnected the music industry is and which artists play central roles in shaping collaboration trends.



